The Unicode Cookbook \linebreak for Linguists

### 2018-07-31 book/chapters/ipa_background.tex
adds backslashes after eg
corrections chapter 4 IPA

In this chapter we present a brief history of the IPA 
(Section~\ref{IPAhistory}), which dates 
back to the late 19th century, not long after the creation of the first 
typewriter with a QWERTY 
keyboard. An understanding of the IPA and its premises and principles 
(Section~\ref{IPApremises-principles}) leads to a better 
appreciation of the challenges that the International Phonetic Association 
faced when digitally encoding the IPA's set of symbols and 
diacritics (Section~\ref{EncodingIPA}). Occurring a little over a hundred years after 
the inception of the IPA, its encoding was a major challenge 
(Section~\ref{need-for-multilingual-environment}); many 
linguists have encountered the pitfalls when the two are used together 
(Chapter~\ref{ipa-meets-unicode}).

@@

Over the years there have been several revisions, but mostly minor ones. Articulation 
labels -- what are often called \textit{features} even though the IPA
deliberately avoids this term -- have changed, e.g.\ terms like \textit{lips}, \textit{throat}
or \textit{rolled} are no longer used. Phonetic symbol values have changed, e.g.\
voiceless is no longer marked by <h>. Symbols have been dropped, e.g.\ the
caret diacritic denoting `long and narrow' is no longer used. And many symbols
have been added to reflect contrastive sounds found in the world's very diverse
phonological systems. The use of the IPA is guided by principles outlined in 
the \textit{Handbook of the International Phonetic Association} \citep{IPA1999}, 
henceforth simply called \textit{Handbook}.

@@

In 1989, an IPA revision convention was held in Kiel, Germany. As in previous meetings, 
there were changes made to the repertoire of phonetic symbols in the IPA chart, which 
reflected what had been discovered, described and cataloged by linguists about the 
phonological systems in the world's languages in the interim. Personal computers 
were also becoming more commonplace, and linguists were using them to create databases. 
A cogent example is the UCLA Phonological 
Segment Inventory Database (UPSID; \citealt{Maddieson1984}), which was expanded 
\citep{MaddiesonPrecoda1990} and then encoded and distributed in a computer program 
\citep{MaddiesonPrecoda1992}.\footnote{It could be installed via 
floppy disk on an IBM PC, or compatible, running 
DOS with 1MB free disk space and 360K available RAM.} The programmers used 
only ASCII characters to maximize compatibility (e.g.\ <kpW> for $[$kpʷ$]$), but 
were faced with unavoidable arbitrary mappings between ASCII letters and 
punctuation, and the more than 900 segment types documented 
in their sample of world's languages' phonological systems. The developers 
devised a system of base characters with secondary diacritic marks 
(e.g.\ in the previous example <kp>, the base character, is modified with <W>). 
This encoding approach is 
also used in SAMPA and X-SAMPA (Section~\ref{sampa-xsampa}) and in the 
ASJP.\footnote{See the ASJP use case in the online supplementary 
materials to this book: \url{https://github.com/unicode-cookbook/recipes}.} 
But before UPSID, SAMPA and ASJP, IPA was encoded with numbers.

@@

% "The set of IPA symbols and their numbers were used to draw up an entity set within SGML by the Text Encoding Initiative (TEl). The name of each entity is formed by 'IPA' preceding the number, e.g. IPA304 is the rEIentity name of lower-case A. These symbols can be processed as IPA symbols and represented on paper and screen with the appropriate local font by modifying the :entity replacement text. The advantage of the SGML entity set is that it is independent or the character set being used."
% "A TEl writing system declaration (wsd) has been drawn up for the IPA symbols."
% A TEl writing system declaration (wsd) has been drawn up for the IPA symbols. This document gives information about the symbol and its IPA function, as well as its encoding in the accompanying SGML document and in UnicodelUCS and in AFII. The writing system declaration can be read as a text d9cument or processed by machines in an SMGL process.

>>>>

In this chapter we present a brief history of the IPA 
(Section~\ref{IPAhistory}), which dates 
back to the late 19th century, not long after the creation of the first 
typewriter with a QWERTY 
keyboard. An understanding of the IPA and its premises and principles 
(Section~\ref{IPApremises-principles}) leads to a better 
appreciation of the challenges that the International Phonetic Association 
faced when digitally encoding the IPA's set of symbols and 
diacritics (Section~\ref{EncodingIPA}). Occurring a little over a hundred years after 
the inception of the IPA, its encoding was a major challenge 
(Section~\ref{need-for-multilingual-environment}); many 
linguists have encountered pitfalls when the two are used together 
(Chapter~\ref{ipa-meets-unicode}).

@@

Over the years there have been several revisions, but mostly minor ones. Articulation 
labels -- what are often called \textit{features}, even though the IPA
deliberately avoids this term -- have changed, e.g.\ terms like \textit{lips}, \textit{throat}
or \textit{rolled} are no longer used. Phonetic symbol values have changed, e.g.\
voiceless is no longer marked by <h>. Symbols have been dropped, e.g.\ the
caret diacritic denoting `long and narrow' is no longer used. And many symbols
have been added to reflect contrastive sounds found in the world's very diverse
phonological systems. The use of the IPA is guided by principles outlined in 
the \textit{Handbook of the International Phonetic Association} \citep{IPA1999}, 
henceforth simply called \textit{Handbook}.

@@

In 1989, an IPA revision convention was held in Kiel, Germany. As in previous meetings, 
there were changes made to the repertoire of phonetic symbols in the IPA chart, which 
reflected what had been discovered, described and cataloged by linguists about the 
phonological systems in the world's languages in the interim. Personal computers 
were also becoming more commonplace, and linguists were using them to create databases. 
A cogent example is the UCLA Phonological 
Segment Inventory Database (UPSID; \citealt{Maddieson1984}), which was expanded 
\citep{MaddiesonPrecoda1990} and then encoded and distributed in a computer program 
\citep{MaddiesonPrecoda1992}.\footnote{It could be installed via 
floppy disk on an IBM PC, or compatible, running 
DOS with 1MB free disk space and 360K available RAM.} The programmers used 
only ASCII characters to maximize compatibility (e.g.\ <kpW> for $[$kpʷ$]$), but 
were faced with unavoidable arbitrary mappings between ASCII letters and 
punctuation, and the more than 900 segment types documented 
in their sample of world's languages' phonological systems. The developers 
devised a system of base characters with secondary diacritic marks 
(e.g.\ in the previous example <kp>, the base character, is modified with <W>). 
This encoding approach is 
also used in SAMPA and X-SAMPA (see below) and in the 
ASJP.\footnote{See the ASJP use case in the online supplementary 
materials to this book: \url{https://github.com/unicode-cookbook/recipes}.} 
But before UPSID, SAMPA and ASJP, IPA was encoded with numbers.

@@

% "The set of IPA symbols and their numbers were used to draw up an entity set within SGML by the Text Encoding Initiative (TEl). The name of each entity is formed by 'IPA' preceding the number, e.g.\ IPA304 is the rEIentity name of lower-case A. These symbols can be processed as IPA symbols and represented on paper and screen with the appropriate local font by modifying the :entity replacement text. The advantage of the SGML entity set is that it is independent or the character set being used."
% "A TEl writing system declaration (wsd) has been drawn up for the IPA symbols."
% A TEl writing system declaration (wsd) has been drawn up for the IPA symbols. This document gives information about the symbol and its IPA function, as well as its encoding in the accompanying SGML document and in UnicodelUCS and in AFII. The writing system declaration can be read as a text d9cument or processed by machines in an SMGL process.
